---
title: "02-ice-data"
output: html_document
date: "2025-06-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rvest)
library(tidyverse)
library(R.utils)
library(raster)
library(sf)
```

## Data Acquisition

Download and access the US National ice cover dataset (NOAA dataset ID G02156) at a 1km resolution. This dataset has daily measurements and covers the northern hemisphere including Fogo island and Newfoundland. 

I'll look at number of days with snow cover over the period of November to May, and the frequency of changes between snow and uncovered?

first let's download the data products adn crop to the island of fogo.
```{r ice download, message=FALSE, results='hide'}
ice_dir <- "~/development/phd-repositories/antler/data/icedata"
years_we_want <- paste("20", seq(from = 16, to = 25), sep = "")


for (year in years_we_want) {
  dir.create(file.path(ice_dir, year), showWarnings = FALSE) #warnings silenced in case directory exists already
  setwd(file.path(ice_dir, year)) # will only be valid for this chunk in rmd files
  
  #download all the files for each year
  url <- paste("https://noaadata.apps.nsidc.org/NOAA/G02156/GIS/1km/", year, "/", sep = "")
  page <- read_html(url)
  files <- page %>% 
    html_nodes("a") %>% 
    html_attr("href")
  
  #for cropping the whole image to just fogo, only needs to be defined once
  box_path <- "/Users/legs/development/phd-repositories/antler/data/icedata/fogo_bounding_box.gpkg"
  fogo_box <- st_read(box_path,
                      layer = "fogo_box")
  
  for(i in 2:length(files)){ 
    u <- paste(url,files[i], sep="/")  
    download.file(u,files[i], mode = "wb")}
  
  # cleanup and process the files
  files_in_dir <- list.files(file.path(ice_dir, year)) # list all downloaded files
  files_df <- data.frame(file_name = files_in_dir) %>% # create a df with file names and months for iteration
    mutate(file_date = str_extract(file_name, "(?<=ims)\\d{7}"), 
           date_obj = as.Date(file_date, format = "%Y%j"),
           month = month(date_obj),
           month_names = as.character(month(date_obj, label = TRUE))) %>% 
    filter(month %in% c(1:5, 11:12))
  
  #remove files that aren't in the date range we're interested in (i.e., remove june to october)
  files_to_remove <- list(files_in_dir[!(files_in_dir %in% files_df$file_name)])
  do.call(unlink, files_to_remove)
  
  for (month_name in unique(files_df$month_names)) {
    this_month <- files_df %>% 
    filter(month_names == month_name) %>% 
    pull(file_name)
    dir.create(file.path(ice_dir, year, month_name), showWarnings = FALSE)
    
    for (file in this_month) {
      gunzip(file.path(ice_dir, year, file), 
             destname = file.path(ice_dir, year, month_name, paste(gsub("[.]gz$", "", file))), remove = TRUE) 
      # Note: you could change to remove = TRUE and it will delete the .gz file, but I will manually delete
    }
    for (tif_file in list.files(file.path(ice_dir, year, month_name))) {
      raster <- raster(file.path(ice_dir, year, month_name, tif_file))
      cropped_raster <- crop(raster, fogo_box)
      writeRaster(cropped_raster, file.path(ice_dir, year, month_name, tif_file), format = "GTiff", overwrite = TRUE)
    }
  }
}
```


## Data Analysis

stack rasters to get summary stats from them, average days with snow cover, # of freeze/cover and thaw events, date snow cover ends?

From inspecting the tifs: *double check these though*
4 = snow/ice cover on land
3 = ice (water)
2 = land without snow
1 = water without ice

*below is untested*
was looking at this link: https://datacarpentry.github.io/r-raster-vector-geospatial/12-time-series-raster.html

potential stats:
using daily percent covered (number of cells with snow/ total cells) I can isolate days/months with no snow (0.0) or total coverage (1.0)
how much did each cell fluctuate?
```{r raster prep}

snow_cover <- data.frame(matrix(ncol=6,nrow=0, dimnames=list(NULL, c("file_name", "avg_snow_cover", "date", "day", "month", "year"))))
#repeat for each combination (and join) after finalizing stats I want to pull out
for (this_year in years_we_want) {
  for (this_month in list.files(path = file.path(ice_dir, this_year))) { #only select months we have data for
    # read in rasters, stack, and make logical (land covered in snow or not)
    rastlist <- list.files(path = file.path(ice_dir, this_year, this_month), full.names= T)
    rstack <- raster::stack(rastlist)
    logical_stack <- calc(rstack, fun = function(x) {
      ifelse(x == 4, TRUE, # true is 1, false is 0, NA for water codes
         ifelse(x == 2, FALSE, NA))
    })
    
    #generate average snow cover on fogo
    year_month_df <- data.frame(avg_snow_cover = cellStats(logical_stack, stat = 'mean')) %>%  
      rownames_to_column(., var = "file_name") %>% 
      mutate(date = as.Date(str_extract(file_name, "(?<=ims)\\d{7}"), format = "%Y%j"), 
             day = format(date, "%d"), 
             month = format(date, "%m"), 
             year = format(date, "%Y"))
    #and add to larger data.frame
    snow_cover <- rbind(snow_cover, year_month_df)
  }
}
```

plotting snow cover data

```{r snow plots}
snow_cover_plot <- snow_cover %>% 
  mutate(calving_season = as.factor(ifelse(snow_cover$month >= 11,
                                             paste(snow_cover$year, as.numeric(snow_cover$year) + 1, sep="/"),
                                             paste(as.numeric(snow_cover$year) - 1, snow_cover$year, sep="/")))) %>% 
  group_by(calving_season) %>%
  arrange(date) %>%
  mutate(day_of_season = row_number()) %>%
  ungroup() %>% 
  filter(!calving_season %in% c("2019/2020", "2020/2021", "2015/2016"))


ggplot(snow_cover_plot, aes(x = day_of_season, y = avg_snow_cover, color = calving_season)) + 
  ylim(0, NA) +
  geom_smooth( se = FALSE) +
  scale_color_manual(values = c("2024/2025" = "tomato3", "2023/2024" = "tomato", "2022/2023" = "royalblue3", "2021/2022" = "dodgerblue3", "2018/2019" = "cornflowerblue", "2017/2018" = "skyblue2", "2016/2017" = "lightblue2"),  name = "Calving Season") +
  scale_x_continuous(breaks = c(1, 31, 62, 93, 121, 151, 182), labels = c('N', 'D', 'J', 'F', 'M', 'A', 'M')) +
  labs(x = "Month", y = "Mean Snow Cover") +
  theme_minimal(base_size = 24)

snow_cover_monthly <- snow_cover_plot %>% 
  group_by(calving_season, month) %>% 
  summarize( max_cover = max(avg_snow_cover), min_cover = min(avg_snow_cover), avg_cover_monthly = mean(avg_snow_cover)) %>% 
  arrange(month) %>% 
  mutate(month_of_season = as.factor(rep(c(3:7,1,2))))

ggplot(snow_cover_monthly, aes(x = month_of_season, y = avg_cover_monthly, color = month_of_season)) + 
  geom_boxplot() +
  scale_x_discrete(labels = c('N', 'D', 'J', 'F', 'M', 'A', 'M'))

ggplot(snow_cover_monthly, aes(x = month_of_season, y = avg_cover_monthly, color = calving_season, group = calving_season)) + 
  geom_line() +
  scale_x_discrete(labels = c('N', 'D', 'J', 'F', 'M', 'A', 'M'))
```


