---
title: "Temperature Analysis"
output: html_notebook
---

R download climate data from ECCC

Need data from 2016-2024 for Twillingate AUT (Station ID = 7083, Climate ID = 8404025), hourly/daily/monthly data available
note: Twillingate station (a separate station from Twilingate AUT) is not in use (was in the 1960s)

To gather the publicly available weather data
```{r twillingate data, eval = false}
library(weathercan)
stations_dl()
twillingate <- weather_dl(station_ids = 7083, start = "2016-01-01", end = "2024-12-31")

#write.csv(twillingate, "twillingate_weather.csv")
```


## Data exploration
visualize the weather changes per year
```{r data prep, include = FALSE}
library(tidyverse)
library(mgcv)
library(ggplot2)
library(nlme)
library(gratia)
library(flextable)

twillingate <- read.csv('../data/twillingate_weather.csv')
```


## Gamm Analysis
Fitting a gamm to this data to investigate changes between years. Focus on 2023/2024 calving season where we observed abnormal antler growth timing.

```{r gamm data prep}
# Load the dataset (Ensure the file path is correct)
calving_season_data <- read.csv("twillingate_weather.csv")

# Ensure date columns are properly formatted
calving_season_data$date <- as.Date(with(calving_season_data, paste(year, month, day, sep="-")), "%Y-%m-%d")

# Assign calving season variable
calving_season_data$calving_season <- ifelse(calving_season_data$month >= 11,
                                             paste(calving_season_data$year, calving_season_data$year + 1, sep="/"),
                                             paste(calving_season_data$year - 1, calving_season_data$year, sep="/"))

# Convert calving season to a factor and set 2022/2023 as the reference level
calving_season_data$calving_season <- factor(calving_season_data$calving_season, 
                                             levels = c("2023/2024", sort(unique(calving_season_data$calving_season[calving_season_data$calving_season != "2023/2024"]), decreasing = T)))

# Filter for relevant months (Nov–May)
calving_season_data <- subset(calving_season_data, month %in% c(11, 12, 1, 2, 3, 4, 5))

# Aggregate data to daily values with mean, min, and max temperature
calving_season_daily <- calving_season_data %>%
  group_by(date, calving_season) %>%
  summarise(mean_temp = mean(temp, na.rm = TRUE),
            min_temp = min(temp, na.rm = TRUE),
            max_temp = max(temp, na.rm = TRUE),
            precip_amt = sum(precip_amt, na.rm = TRUE),
            .groups = 'drop') %>% 
  filter(!calving_season %in% c("2015/2016", "2024/2025")) #remove incomplete seasons

# Create day of season variable
calving_season_daily <- calving_season_daily %>%
  group_by(calving_season) %>%
  arrange(date) %>%
  mutate(day_of_season = row_number()) %>%
  ungroup()

#create a trimmed dataset for a more targetted model
calving_season_daily_trim <- calving_season_daily %>% 
  mutate(month = format(date, "%m")) %>% 
  filter(month %in% c("02","03","04"))  

# Create day of season variable
calving_season_daily_trim <- calving_season_daily_trim %>%
  group_by(calving_season) %>%
  arrange(date) %>%
  mutate(day_of_season = row_number()) %>% #now 1 is February 1st
  ungroup()

```

### Gamm for Winter months (November to end of May)
Fit a gamm model with an interaction between year (calving_season) and time throughout the winter (day_of_season, count of days starting nov 1 to may 31)
```{r Gamm temp}
# Fit a Generalized Additive Mixed Model (GAMM) with AR(1) autocorrelation within calving season
gam_model <- gamm(mean_temp ~ s(day_of_season, by=calving_season) + calving_season, 
                  data = calving_season_daily, 
                  family = gaussian(), 
                  correlation = corAR1(form = ~ day_of_season | calving_season), 
                  method = "REML")

# Summary of the model
summary(gam_model$gam)
```

#### Gamm visualizations

Most years are significantly cooler than abnormal year (coloured red), but specifically over the mid winter months and early spring (feb to end of april) this year is warmer. It's cooler earlier in the winter/late fall though.
```{r visualizations}
# Post-hoc test to compare smooth trends across calving seasons
sm_diff <- difference_smooths(gam_model$gam, select = "s(day_of_season)", ci = 0.95)
sm_diff <- sm_diff %>% 
  filter(.level_1 == "2023/2024" | .level_2 == "2023/2024") %>% #only relevant comparisons
  filter(!.level_1 %in% c("2019/2020", "2020/2021") & !.level_2 %in% c("2019/2020", "2020/2021")) #maybe colour code the years as sig or not?
draw(sm_diff)
# Visualization of modeled temperature trends across calving seasons
plot_data <- expand.grid(
  day_of_season = seq(1:212),
  calving_season = unique(calving_season_daily$calving_season),
  precip_amt = mean(calving_season_daily$precip_amt, na.rm = TRUE))

plot_data <- plot_data %>% 
  mutate(predicted_temp = 
           predict(gam_model$gam, newdata = plot_data, type = "response")) %>% 
  filter(!calving_season %in% c("2019/2020", "2020/2021"))

gamm_plot <- ggplot(calving_season_daily, aes(x = day_of_season, y = mean_temp, color = calving_season)) +
  scale_color_manual(values = c("2023/2024" = "tomato", "2022/2023" = "royalblue3", "2021/2022" = "dodgerblue3", "2018/2019" = "cornflowerblue", "2017/2018" = "skyblue2", "2016/2017" = "lightblue2"),  name = "Calving Season") +
  #geom_point(alpha = 0.3) +
  geom_line(data = plot_data, aes(x = day_of_season, y = predicted_temp), size = 1) +
  scale_x_continuous(breaks = c(1, 31, 62, 93, 121, 151, 182), labels = c('N', 'D', 'J', 'F', 'M', 'A', 'M')) +
  labs(x = "Month", y = "Mean Temperature (°C)", ) +
  theme_minimal(base_size = 24)

gamm_plot
ggsave('../plots/gamm_plot_all_winter.png',
       gamm_plot,
       width = 10,
       height = 5,
       dpi = 320
    )

```


```{r summary table for publication}
library(flextable)

table_s2 <- as_flextable(gam_model$gam)
save_as_docx(table_s2, path = "../plots/table_s2.docx")
```

### Narrower date window (Feb to end of April)
If we narrow the window to only february-april? These are the months that appear warmer from the first gamm.
All are significantly cooler than abnormal year in this time period.

Day of season has been changed to a count that starts February 1st.
```{r targeted gamm}
# Fit a Generalized Additive Mixed Model (GAMM) with AR(1) autocorrelation within calving season
gam_model_trim <- gamm(mean_temp ~ s(day_of_season, by=calving_season) + calving_season, 
                  data = calving_season_daily_trim, 
                  family = gaussian(), 
                  correlation = corAR1(form = ~ day_of_season | calving_season), 
                  method = "REML")

# Summary of the model
summary(gam_model_trim$gam)
```


#### Visualizations of trimmed gamm. 
The abnormal year is almost linear over this time period. I'm not sure we want to use this at all but now it exists.
```{r trimm gamm visuals}
# Post-hoc test to compare smooth trends across calving seasons
sm_diff_trim <- difference_smooths(gam_model_trim$gam, select = "s(day_of_season)", ci = 0.95)
sm_diff_trim <- sm_diff_trim %>% 
  filter(.level_1 == "2023/2024" | .level_2 == "2023/2024") %>% #only relevant comparisons
  filter(!.level_1 %in% c("2019/2020", "2020/2021") & !.level_2 %in% c("2019/2020", "2020/2021")) #maybe colour code the years as sig or not?
draw(sm_diff_trim)
# Visualization of modeled temperature trends across calving seasons
plot_data_trim <- expand.grid(
  day_of_season = seq(1:90),
  calving_season = unique(calving_season_daily_trim$calving_season),
  precip_amt = mean(calving_season_daily_trim$precip_amt, na.rm = TRUE))

plot_data_trim <- plot_data_trim %>% 
  mutate(predicted_temp = 
           predict(gam_model_trim$gam, newdata = plot_data_trim, type = "response")) %>% 
  filter(!calving_season %in% c("2019/2020", "2020/2021"))

gamm_plot_trim <- ggplot(calving_season_daily_trim, aes(x = day_of_season, y = mean_temp, color = calving_season)) +
  scale_color_manual(values = c("2023/2024" = "tomato", "2022/2023" = "royalblue3", "2021/2022" = "dodgerblue3", "2018/2019" = "cornflowerblue", "2017/2018" = "skyblue2", "2016/2017" = "lightblue2")) +
  #geom_point(alpha = 0.3) +
  geom_line(data = plot_data_trim, aes(x = day_of_season, y = predicted_temp), size = 1) +
  scale_x_continuous(breaks = c(0, 28, 59, 89), labels = c( 'Feb', 'Mar', 'Apr', 'May')) +
  labs(x = "Month", y = "Mean Temperature (°C)",
       title = "Modeled Temperature Trends Across Calving Seasons") +
  theme_minimal()

gamm_plot_trim

ggsave('../plots/gamm_plot_trimmed.png',
       gamm_plot_trim,
       width = 10,
       height = 7,
       dpi = 320
    )
```

```{r gamm table trimmed}

table_s3 <- as_flextable(gam_model_trim$gam)
save_as_docx(table_s3, path = "../plots/table_s3.docx")
```

